Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-07-05 _tex/paper2.tex
Built site for gh-pages

\usepackage{float}
\floatplacement{table}{H}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Psychometrika}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, directed acyclic graphs, structural
causal models, bayesian statistical methods, thurstonian
model, comparative judgement, probability, statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

As in the previous section, we begin defining the \emph{comparison
mechanism} using the binary vector variable \(C\) to facilitate the
interpretation of the sample-comparison model. Equation (\ref{eq-mat27})
shows that \(C\) contains \(n\) elements corresponding to the number of
rows in the \(R\) matrix, with each element \(c_{(i,a,h,b,j,k)}\) being
a binary value indicating the presence or absence of data rows in \(R\),
a definition similar to that of \(s_{(i,a,h,b,j,k)}\) in Equation
(\ref{eq-mat26}). \begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

@@

\caption{\label{fig-cj15}Comparative judgment model}

@@

\caption{\label{fig-cj16}Comparative judgment model, SCM, probabilistic
and statistical model assuming different discriminal dispersions for the
student's traits}

@@

This formulation reveals that the discriminal difference captures a
\emph{pure interaction effect}, in which neither the texts' discriminal
processes nor the judges' biases alone determine the outcome, but their
interaction does \citep{Attia_et_al_2022}. Put simply, this mathematical
description captures the idea that the stimuli' discriminal processes
become an observable outcome only through the lens of judges'
perceptions (i.e., their biases). For clarity, the square brackets in
\(D_{R}\) indicate the relevant indices for each trait vector; they do
not imply any subsetting of the data.

@@

The \textbf{Declarations} section of this document provides a link to
the model code, along with an alternative specification that assumes
equal discriminal dispersions. We tested both versions of the model with
success using \texttt{Stan} \citep[version 2.26.1]{Stan_2020}.

@@

Applying the approach effectively requires foundational knowledge in two
areas. First, a solid understanding of causal inference principles.
Second, the ability to translate the functional and probabilistic
aspects of conceptual models into bespoke statistical models. A clear
example illustrating the importance of causal inference is the recurrent
assumption that predictor variables are ``relevant'' to the research
context--interpreting this relevance as their inclusion in a
\emph{sufficient adjustment set}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, this
study does not explore the full implications of that assumption for
model specification, estimation, and inference. Nevertheless, to assist
with the effort of engaging with this and other complex causal inference
concepts, key references are provided throughout the study to guide
applied CJ researchers toward a deeper understanding of these ideas
\footnote{Also refer to the detailed online document referenced in the
  Declarations section}.

@@

Even after acquiring the necessary foundational knowledge, applying this
approach to a specific CJ context involves two additional challenges.
First, it is essential to verify whether the conceptual model offers a
\emph{faithful} \footnote{Avoid confusing this term with the
  \emph{faithfulness assumption} \citep{Neal_2020, Hernan_et_al_2025}
  described in the causal inference literature. For further details, see
  footnote 1.} representation of the CJ system under study. Second, the
statistical translation of the model must be assessed to determine if it
can accurately estimate the intended estimands (i.e., parameters) from
empirical data. To ensure fidelity, the conceptual and statistical
models presented here should be treated as starting points, not
universal solutions for all CJ designs or datasets. While these models
may be adequate in some contexts, assuming so without evaluation is
unwise. Instead, adaptation to the specific context is necessary, with
careful attention to assessment design features and causal assumptions.
As discussed in Section~\ref{sec-discussion_challenges}, this approach
supports such adaptation by providing a transparent framework for
articulating new assumptions and guiding CJ assessment design.

>>>>

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Psychometrika}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, directed acyclic graphs, structural
causal models, bayesian statistical methods, thurstonian
model, comparative judgement, probability, statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

As in the previous section, we begin defining the \emph{comparison
mechanism} using the binary vector variable \(C\) to facilitate the
interpretation of the CSM. Equation (\ref{eq-mat27}) shows that \(C\)
contains \(n\) elements corresponding to the number of rows in the \(R\)
matrix, with each element \(c_{(i,a,h,b,j,k)}\) being a binary value
indicating the presence or absence of data rows in \(R\), a definition
similar to that of \(s_{(i,a,h,b,j,k)}\) in Equation (\ref{eq-mat26}).
\begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

@@

\caption{\label{fig-cj15}Comparative judgment model.}

@@

\caption{\label{fig-cj16}Comparative judgment model. SCM, probabilistic
and statistical model assuming different discriminal dispersions for the
student's traits}

@@

This formulation reveals that the discriminal difference captures a
\emph{pure interaction effect}, in which neither the texts' discriminal
processes nor the judges' biases alone determine the outcome, but their
interaction does \citep{Attia_et_al_2022}. Put simply, this mathematical
description captures the idea that the stimuli' discriminal processes
become an observable outcome only through the lens of judges'
perceptions (i.e., their biases). For clarity, the square brackets in
\(D_{R}\) indicate the relevant indices for each trait vector.

@@

The \textbf{Declarations} section of this document provides a link to
the model code, along with an alternative specification that assumes
equal discriminal dispersions. Both versions of the model have been
tested with success using \texttt{Stan} \citep[version
2.26.1]{Stan_2020}.

@@

Applying the approach effectively requires foundational knowledge in two
areas. First, a solid understanding of causal inference principles.
Second, the ability to translate the functional and probabilistic
aspects of conceptual models into bespoke statistical models. A clear
example illustrating the importance of causal inference is the recurrent
assumption that predictor variables are ``relevant'' to the research
context--interpreting this relevance as their inclusion in a
\emph{sufficient adjustment set}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, this
study does not explore the full implications of that assumption for
model specification, estimation, and inference. Nevertheless, to assist
with the effort of engaging with this and other complex causal inference
concepts, key references are provided throughout the study to guide
applied CJ researchers toward a deeper understanding of these ideas
\footnote{Also refer to the detailed online document referenced in the
  \textbf{Declarations} section}.

@@

Even after acquiring the necessary foundational knowledge, applying this
approach to a specific CJ context involves two additional challenges.
First, it is essential to verify whether the conceptual model achieves
sufficient \emph{fidelity} in representing the CJ system under study.
Second, the statistical translation of the model must be assessed to
determine if it can accurately estimate the intended estimands (i.e.,
parameters) from empirical data. To ensure fidelity, the conceptual and
statistical models presented here should be treated as starting points,
not universal solutions for all CJ designs or datasets. While these
models may be adequate in some contexts, assuming so without evaluation
is unwise. Instead, adaptation to the specific context is necessary,
with careful attention to assessment design features and their implied
causal assumptions. As discussed in
Section~\ref{sec-discussion_challenges}, this approach supports such
adaptation by providing a transparent framework for articulating new
assumptions and guiding CJ assessment design.
